---
title: "The 4 parameter logistic"
description: |
  Growth and change between a minimum and a maximum.
author:
  - name: Andrew and Will
    url: {}
date: 06-28-2022
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
library(cmdstanr)
library(tidybayes)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

I want to make a simple and simulated version of Will's logistic model. 

thinking about treating L_o on the log scale -- ie the log of a number between 0 and 1

```{r}
curve(log(x))
```

$$
\begin{align}
y &= L + \frac{1 - L}{1 + e^{-\beta}} \\ 
 &= L + (1 - L) \times \frac{1}{1 + e^{-\beta}}\\
\text{log}(y) &= \text{log}\left(L + (1-L)\times \frac{1}{1 + e^{-\beta}}\right)
\end{align}
$$

Return the log mixture of the log densities lp1 and lp2 with mixing proportion theta, defined by \[\begin{eqnarray*} \mathrm{log\_mix}(\theta, \lambda_1, \lambda_2) & = & \log \!\left( \theta \exp(\lambda_1) + \left( 1 - \theta \right) \exp(\lambda_2) \right) \\[3pt] & = & \mathrm{log\_sum\_exp}\!\left(\log(\theta) + \lambda_1, \ \log(1 - \theta) + \lambda_2\right). \end{eqnarray*}\]
Available since 2.6



```{r}


log_mix_logistic <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/log_mix_logistic.stan"))

datlist <- list(n = 50,
                x = runif(50, min = -40, max = 40))

demo_samples <- log_mix_logistic$sample(data = datlist, refresh = 0)



```

Very VERY interesting note -- if you do this in `tranformed parameters` it doesn't work at all! 

```{r}


demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y = exp(.value), group = .draw)) + geom_line()

```

```{r}

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y = .value, group = .draw)) + geom_line()

```


```{r}

log_sum_exp_logistic <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/log_sum_exp_logistic.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- log_sum_exp_logistic$sample(data = datlist, refresh = 0)
```


```{r}


demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y = exp(.value), group = .draw)) + geom_line()
```

could the expression be vectorized easily? does that matter?

```{r}

vec_logistic <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/vec_logistic.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- vec_logistic$sample(data = datlist, refresh = 0)
```

```{r}
demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y =.value, group = .draw)) + geom_line()

```

prior simulations in R 

```{r}
x <-  seq(from = -40, to = 40, length.out = 50)
L <- rnorm(1, -1, 1)
alpha <- exp(rnorm(1, -.8, 1))
y <- plogis(L) + (1 - plogis(L))*plogis(alpha*x)
plot(x, y, ylim = c(0, 1))
```

```{r}
over_vectorized <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/over_vectorized.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- over_vectorized$sample(data = datlist, refresh = 0)
```


```{r}
demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y =exp(.value), group = .draw)) + geom_line()

```

### adding a maximum

Here I'm adding another parameter to control the (log) of the maximum:

```{r}
vec_logistic_max <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/vec_logistic_max.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- vec_logistic_max$sample(data = datlist, refresh = 0)
```


```{r}
demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y =.value, group = .draw)) + geom_line()

```

That looks about right! We can take the same approach with the approach using `log_sum_exp`




```{r}
log_sum_exp_logistic_max <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/log_sum_exp_logistic_max.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- log_sum_exp_logistic_max$sample(data = datlist, refresh = 0)
```


```{r}
demo_draws <- gather_draws(demo_samples, y[i], ndraws = 50)

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y =exp(.value), group = .draw)) + geom_line()

```

### simulate lognormal data

I'd like to try to tack on a logistic simulation to each of these, just to see!



```{r}
log_sum_exp_ln_rng <- cmdstanr::cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/log_sum_exp_ln_rng.stan"))

datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50))

demo_samples <- log_sum_exp_ln_rng$sample(data = datlist, refresh = 0)
```


```{r}
get_variables(demo_samples)
demo_dots <- gather_draws(demo_samples, y_obs[i], ndraws = 5)

demo_dots |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, y = .value, group = .draw)) + geom_point() + 
  facet_wrap(~.draw)

```

could also draw the line ribbon

```{r}

demo_draws <- gather_rvars(demo_samples, y_obs[i])

demo_draws |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, dist = .value)) + stat_dist_lineribbon()

```

Now try to fit it, why not!?

```{r}
log_sum_exp_ln_model <- cmdstan_model(stan_file = here::here("_posts/2022-06-28-nonlinear-functions-for-growth/log_sum_exp_ln_model.stan"))


datlist <- list(n = 50,
                x = seq(from = -40, to = 40, length.out = 50),
                yy = demo_dots |> filter(.draw == max(.draw)) |> pull(.value)
                )
model_samples <- log_sum_exp_ln_model$sample(data = datlist, refresh = 0)

model_samples
```

```{r}

model_mean <- gather_rvars(model_samples, log_mu[i])

model_mean |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, dist = exp(.value))) + 
  stat_dist_lineribbon() + 
  geom_point(aes(x = x, y = yy), inherit.aes = FALSE,
             data = as.data.frame(datlist))
```

```{r}

model_pred <- gather_rvars(model_samples, y_obs[i])

model_pred |> 
  ungroup() |> 
  mutate(x = datlist$x[i]) |> 
  ggplot(aes(x = x, dist = .value)) + 
  stat_dist_lineribbon() + 
  geom_point(aes(x = x, y = yy), inherit.aes = FALSE,
             data = as.data.frame(datlist), pch = 21, fill = "orange") + 
  scale_fill_brewer(palette = "Greens")
```

