---
title: "Week6"
description: |
  A short description of the post.
author:
  - name: Andrew
    url: {}
date: 2022-07-22
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
suppressPackageStartupMessages(library(rethinking))

data("reedfrogs")

head(reedfrogs)
```

using the model described [here](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-12?slide=36)

```{r}
unique(reedfrogs$density)
```

```{r}
library(tidyverse)


logit_average_survival <- rnorm(1, mean = 0, sd = 1)
logit_sd_survival <- rexp(1, rate = 1)

reedfrogs |> 
  select(density) |> 
  mutate(logit_tank_surv = rnorm(length(density), mean = logit_average_survival, sd = logit_sd_survival))
```

## lil simulation

```{r}
data.frame(y = letters) |> 
  mutate(ff = rnorm(1))
```



```{r}
library(cmdstanr)

trt_effects <- cmdstan_model(stan_file = here::here("_posts/2022-07-22-week6/random_effects.stan"), pedantic = TRUE)

```



```{r}
d <- reedfrogs
dat <- list(
S = d$surv,
D = d$density,
T = 1:nrow(d),
P = ifelse( d$pred=="no" , 1L , 2L ),
G = ifelse( d$size=="small" , 1L , 2L ) )
m2 <- ulam(
alist(
S ~ binomial( D , p ),
logit(p) <- a[T] + b[P,G],
a[T] ~ normal( 0 , sigma ),
matrix[P,G]:b ~ normal( 0 , 1 ),
sigma ~ exponential( 1 )
), data=dat , chains=4 , cores=4 , log_lik=TRUE , sample = FALSE)


stancode(m2)
```

```{r}
reed_data <- with(reedfrogs, 
                  list(
                    n = length(surv),
                    surv = surv,
                    density = density,
                    n_size = nlevels(size),
                    n_pred = nlevels(pred),
                    size_id = as.numeric(size),
                    pred_id = as.numeric(pred)
))

trt_effects$sample(data = reed_data)

```


```{r}

trt_effects <- cmdstan_model(stan_file = here::here("_posts/2022-07-22-week6/treatment_effects.stan"), pedantic = TRUE)
trt_samples <- trt_effects$sample(data = reed_data, refresh = 0)
trt_samples$summary()


```



add back in random effects for tanks

```{r}

tank_trt_effects <- cmdstan_model(stan_file = here::here("_posts/2022-07-22-week6/tank_trt.stan"), pedantic = TRUE)
tank_trt_samples <- tank_trt_effects$sample(data = reed_data, refresh = 0)
tank_trt_samples$summary() |> View()
```




* here we could learn how to "decorate" a model with the factor levels from the data, using [`tidybayes::recover_types`](https://mjskay.github.io/tidybayes/reference/recover_types.html) 

* I used indexing in this model, expecting it to work on arrays as it does in vectors -- and it didn't!

## treatments as random effects

treatments can also be considered random effects! 



```{r}
purrr::walk(1:500, ~ rnorm(10, mean = 0, sd = rexp(1, .5)))
```


### 

```{r}

ordinal_trt_effects <- cmdstan_model(stan_file = here::here("_posts/2022-07-22-week6/ordinal_treat.stan"), pedantic = TRUE)

reed_data <- purrr::list_modify(reed_data,
               dens_id = reedfrogs$density %/% 10,
               n_dens = 3)

ordinal_trt_effects_samples <- ordinal_trt_effects$sample(data = reed_data,
                                                          refresh = 0)
ordinal_trt_effects_samples$summary() |> View()
```

```{r}
rvars <- cbind(
  ordinal_trt_effects_samples |> 
    gather_rvars(pred[h]),
  expand_grid(dens = 1:3, pred = 1:2, size = 1:2) 
)  


probability_plot <- rvars |> 
  ggplot(aes(x = dens, dist = brms::inv_logit_scaled(.value), colour = pred)) + 
  facet_wrap(~size) +
  stat_dist_gradientinterval() + 
  theme_bw()
```

```{r}
reedfwogs <- reedfrogs |> 
  mutate(dens = density %/% 10,
         size = as.numeric(size),
         pred = as.numeric(pred))

probability_plot + 
  geom_point(aes(x = dens, y = propsurv, colour = pred), data = reedfwogs, inherit.aes = FALSE)
```

