---
title: "Estimating the posterior distribution of residuals"
description: |
  A short description of the post.
author:
  - name: Andrew
    url: {}
date: 06-14-2022
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Everything in a Bayesian model has a distribution: there are distributions for the slope coeffiecients, distributions for the residual variance. The model makes a distribution of lines that form the average, and can produce a distribution of predicted values. 

When we measure residuals, therefore, these have distributions too. This is useful because it lets us transmit model uncertainty into these uncertain values.

let's look at bill characteristics of the GEntoo penguin

```{r}
library(palmerpenguins)
suppressPackageStartupMessages(library(tidyverse))
library(patchwork)

gentoo <- penguins |> 
  filter(species == "Gentoo") |> 
  filter(!is.na(bill_depth_mm) & !is.na(bill_length_mm))

depth_fig <- gentoo |> 
  ggplot(aes(x = body_mass_g, y = bill_depth_mm)) + 
  geom_count(alpha = 0.4)

length_fig <- gentoo |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) + 
  geom_count(alpha = .4)

depth_fig + length_fig
```


Not surprisingly, both change with size. let's model this, with a simple gaussian bayesian model

```{r}
library(brms)

len_mass <- brm(bill_length_mm ~ body_mass_g, 
                data = gentoo, refresh = 0,
                backend = "cmdstanr",
                file = here::here("_posts",
                                  "2022-06-14-estimating-the-posterior-distribution-of-residuals",
                                  "len_mass.rds")) 


dep_mass <- brm(bill_depth_mm ~ body_mass_g,
                data = gentoo,
                refresh = 0,
                backend = "cmdstanr",
                file = here::here("_posts",
                                  "2022-06-14-estimating-the-posterior-distribution-of-residuals",
                                  "dep_mass.rds")) 

```

The only essential things here are the model formulae. the arguments `refresh = 0` means the chain progress is not printed to the screen. `backend = cmdstanr` means use `cmdstan` to run the model -- the default `Rstan` would work the same way.

let's look at these modesl

```{r}
len_mass
dep_mass
```


let's draw lines with them

```{r}
library(modelr)
library(tidybayes)

gentoo_lines <- gentoo |> 
  data_grid(body_mass_g = seq_range(body_mass_g, n = 40)) |> 
  add_epred_rvars(len_mass, value = ".epred_len") |> 
  add_epred_rvars(dep_mass, value = ".epred_dep")


dep_epred_fig <- depth_fig + 
  stat_dist_lineribbon(aes(x = body_mass_g, dist = .epred_dep), 
                       data = gentoo_lines, inherit.aes = FALSE) + 
  geom_count(pch = 21, col = "white", fill = "black", alpha = 1)

len_epred_fig <- length_fig + 
  stat_dist_lineribbon(aes(x = body_mass_g, dist = .epred_len), 
                       data = gentoo_lines, inherit.aes = FALSE) + 
  geom_count(pch = 21, col = "white", fill = "black", alpha = 1)

dep_epred_fig + len_epred_fig

```

### Extract residuals

```{r}
gentoo_len_resid <- gentoo |> 
  select(species, bill_length_mm, bill_depth_mm, body_mass_g) |> 
  add_residual_draws(len_mass, value = ".resid_len") |> 
  summarise(.resid_len_mean = mean (.resid_len), 
            .resid_len_sd = sd(.resid_len))

gentoo_dep_resid <- gentoo |> 
  select(species, bill_length_mm, bill_depth_mm, body_mass_g) |> 
  add_residual_draws(dep_mass, value = ".resid_dep") |> 
  summarise(.resid_dep_mean = mean (.resid_dep), 
            .resid_dep_sd    =   sd(.resid_dep))
```

We can visualize thse at the same time

```{r}
gentoo_two_resid <- gentoo_len_resid |> 
  ungroup() |> 
  left_join(ungroup(gentoo_dep_resid), 
            by = c("species", "bill_length_mm", 
                   "bill_depth_mm", "body_mass_g", ".row"))

gentoo_two_resid |> 
  ggplot(aes(x = .resid_len_mean, y = .resid_dep_mean,
             xmin = .resid_len_mean - .resid_len_sd,
             xmax = .resid_len_mean + .resid_len_sd)) + 
  geom_pointrange() + 
  geom_linerange(aes(x = .resid_len_mean, y = .resid_dep_mean,
                     ymin = .resid_dep_mean - .resid_dep_sd,
                     ymax = .resid_dep_mean + .resid_dep_sd))
```


Before moving on, lets demonstrate that the posterior is well-described by a normal distribution

```{r}
one_resid <-  gentoo[5,] |> 
  select(species, bill_length_mm, bill_depth_mm, body_mass_g) |> 
  add_residual_draws(dep_mass, value = ".resid_dep")

mean_draws <-  mean(one_resid$.resid_dep)
sd_draws <- sd(one_resid$.resid_dep)
one_resid |> 
  ggplot(aes(x = .resid_dep)) + 
  geom_histogram(aes(y = ..density..)) + 
  stat_function(fun = function(x) dnorm(x, mean = mean_draws, sd = sd_draws))
```

### error-in-variables regression

```{r}
error_in_var_regression <- brm(.resid_dep_mean | mi(.resid_dep_sd) ~ me(.resid_len_mean, .resid_len_sd),
                               data = gentoo_two_resid,
                               backend = "cmdstanr", file = here::here("_posts",
                                  "2022-06-14-estimating-the-posterior-distribution-of-residuals",
                                  "error_in_var_regression.rds"))
```

