---
title: "Hybrid fly life history"
description: |
  Modelling two parts of the life history of insect development.
author: "Allen and Andrew"
date: 06-06-2022
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---

The goal of this simulation is to simulate the growth and development of insects. 
Two stages of insect development were measured: the time spent as a larva, and time spent as a pupa. We have data on individual flies.

In this experiment, flies come in three kinds: "H", "N", and a hybrid fly "HN"

Flies are developed in batches called "groups", which contain a variable amount of larvae. The members of a group share environments (e.g. the same batch of growing media).

Development time can be variable. However, not every genotype or life history stage will prove equally variable. 
That is, more factors could be relevant to the development of a pupa (which has to melt and reform all its organs) compared to a maggot (a stomach bag with a mouth on one end)

## This example

This is a walkthrough of simulating a Gamma regression. In it we'll discuss:

* the Gamma distribution, and how to cook and eat it.
* Helmert contrasts, aka One Cool Trick for measuring "non additive" effects.
* random effects, both single and multivariate


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(brms)
library(tidyverse)
library(tidybayes)
```

## The gamma distribution

The Gamma distribution is handy for studying continuous positive numbers -- in this case, the number of days in development. There are three ways to write a gamma distribution:

#### The classic

$$
\text{Gamma}(a, b)
$$

The mean $\mu$ and variance $\sigma^2$ is: 
$$
\begin{align}
\mu &= \frac{a}{b} \\
\sigma^2 &= \frac{a}{b^2}
\end{align}
$$

#### The Moment-matcher: mean and sd

You can do some algebra on the expressions above and find out what $a$ and $b$ should be if the mean is $\mu$ and the standard deviation is $\sigma$.
This technique is called Moment Matching.

$$
\text{Gamma}\left(\frac{\mu^2}{\sigma^2}, \frac{\mu}{\sigma^2}\right)
$$

## Mean and rate

$$
\text{Gamma}(\mu b, b)
$$

In this version, the average is $\mu$ and the variance is $\sigma^2 = \mu/b$.  
In other words, the parameter $b$ _describes how the variance scales with the mean._

### Mean and shape

This is the version that you'll find in a lot of stats software, including `brms`. 
It is: 

$$
\text{Gamma}\left(a, \frac{a}{\mu}\right)
$$

Here the mean is once again $\mu$ but the variance is $\sigma^2 = \mu^2/a$

In other words, $a$ _scales the square of the mean to be the variance._ 

<aside>
Is it less confusing to say: $a$ is the square of the number that scales the mean to the standard deviation?
$$
\sigma = \mu/\sqrt{a}
$$
</aside>


see the [brms families vignettes](https://cran.r-project.org/web/packages/brms/vignettes/brms_families.html)

Andrew is going to ask in the Stan forum WTF is happening when you use the gamma family -- why does LF behave differently than the alternative

```{r}
library(tidybayes)
library(brms)
library(tidyverse)
## the Stan and brms parameterization of gamma translates into R like this: 

mean <- 42
shape <- 2

rr <- rgamma(3000, shape, shape/mean)


rr |> mean()
rr |> var()
42^2/2

# equivalently

rr |> sd()
42/sqrt(2)
```

### Recovering these parameters

```{r, eval = FALSE}

gamma_form <- bf(rr ~ 1,
                 family = brmsfamily("Gamma",
                                     link = "log",
                                     link_shape = "log")) + 
  lf(shape ~ 1)

andrew_prior <- c(prior(normal(4, .5), class = "Intercept"),
                  prior(normal(2, .5), class = "Intercept", dpar = "shape"))

gamma_prior_predict <- brm(gamma_form, data = data.frame(rr), 
    prior = andrew_prior,
    backend = "cmdstanr", 
    cores = 4, refresh = 0,
    sample_prior = "yes", 
    file = here::here("_posts", 
                      "2022-06-06-hybrid-fly-life-history",
                      "gamma_prior_predict.rds"),
    file_refit = "on_change")

# get_variables(gamma_prior_predict)

# visualize
gamma_prior_predict |> 
  gather_rvars(prior_Intercept, prior_Intercept_shape, b_Intercept, b_shape_Intercept) |> 
  mutate(is_prior = if_else(condition = str_detect(string = .variable, pattern = "prior.*"),
                            true = "prior", 
                            false = "posterior"),
         param = if_else(condition = str_detect(string = .variable, "shape"),
                         true = "shape", 
                         false = "mean")) |> 
  ggplot(aes(dist = .value, y = param, fill = is_prior)) + 
  stat_dist_halfeye() +
  # coord_cartesian(xlim = c(0,10)) +
  facet_wrap(~param, scales = "free") + 
  geom_vline(aes(xintercept = val), 
             data = tribble(~param, ~val,
                            "mean",  log(mean),
                            "shape", log(shape))
  )
  
```


## experimental design

H, N, and hybrid -- three kinds of flies

is development time different among the three lines? Does the *variance* in development time differ between the lines?


begin by modelling one fly species, one life history stage

using the brms parameterization

```{r}
# larval development times
hist(rgamma(570, shape = 30, rate = 30/12))

# pupal development times
hist(rgamma(570, shape = 30, rate = 30/13))
```

```{r}
gamma_form <- bf(xx ~ 1,
                 family = brmsfamily("Gamma",
                                     link = "log",
                                     link_shape = "log")) + 
  lf(shape ~ 1)

simple_gamma <- brm(gamma_form, 
                    data = data.frame(xx = rgamma(570,
                                                  shape = 30,
                                                  rate = 30/13)),
                    here::here("_posts", 
                               "2022-06-06-hybrid-fly-life-history",
                               "simple_gamma.rds"),
                    backend = "cmdstanr", cores = 4)

simple_gamma |> 
  gather_rvars(Intercept, Intercept_shape) |> 
  mutate(.value = exp(.value)) |> 
  ggplot(aes(y = .variable, dist = .value)) + 
  stat_dist_halfeye()

```


duration and variance are approximately (but not necessarily) equal

Start with: many individuals all with the same expected duration of each life history stage.

correlated both stages

```{r}
corr_larva_pupa <- .8

corr_mat <- matrix(c(1, corr_larva_pupa,
                     corr_larva_pupa, 1), byrow = TRUE, ncol = 2)

larv_sd <- .5
pupa_sd <- .7

sds <- c(larv_sd, pupa_sd)

Sigma <- diag(sds) %*% corr_mat %*% diag(sds)

larv_pupa <- MASS::mvrnorm(570, mu = c(13, 13), Sigma = Sigma)

plot(larv_pupa[,1], larv_pupa[,2])

```

Add observation variance

```{r}

true_shape <- 70

fake_development_times <- tibble(fly_id = 1:570,
       fly_code = paste0("fly", fly_id)) |> 
  mutate(avg_larva = larv_pupa[fly_id, 1],
         avg_pupae = larv_pupa[fly_id, 2]) |> 
  rowwise() |> 
  mutate(
    # add observation variance
    obs_larva = rgamma(1, shape = true_shape, rate = true_shape/avg_larva),
    obs_pupae = rgamma(1, shape = true_shape, rate = true_shape/avg_pupae)
  )

```

plot, then model

```{r}
fake_development_times |> 
  ggplot(aes(x = obs_larva, y = obs_pupae)) + geom_point()


fake_development_times |> 
  ggplot(aes(x = avg_larva, y = avg_pupae)) + geom_point()

```

correlation is hard to see under the gamma errors!


let's see if a model can recover that

```{r}
fake_devo_long <- fake_development_times |> 
  select(-starts_with("avg")) |> 
  pivot_longer(starts_with("obs"),
               names_to = "stage",
               values_to = "days")


gamma_form_correlated_stages <- bf(days ~ stage + (1 + stage | fly_code),
                 family = brmsfamily("Gamma",
                                     link = "log",
                                     link_shape = "log")) + 
  lf(shape ~ 1)


get_prior(gamma_form_correlated_stages, data = fake_devo_long)


corr_stages_priors <- c(
  prior(normal(0,1), class = "b"),
  prior(lkj(2), class = "cor"),
  prior(normal(2.5,1), class = "Intercept"),
  prior(exponential(2), class = "sd"),
  prior(normal(4,2), class = "Intercept", dpar = "shape")
)


correlated_stages_gamma <- brm(gamma_form_correlated_stages,
                               data = fake_devo_long, 
                               prior = corr_stages_priors,
                               file = here::here("_posts", "2022-06-06-hybrid-fly-life-history", "corr_gamma.rds"),
                               cores = 4)

correlated_stages_gamma

```



Maybe a simpler model (the above seems overparameterized) where each larva has a single offset (e.g. proportional to "quality" or "resource use efficiency").

```{r}

nfly <- 570
# average time you spend as a larva
log_avg_larva_time <-  log(13)
# average time DIFFERECE to the time you spend as a pupa (might be close to 0)
log_to_pupa <- -.2
# your "quality", some idea of how much above and below the population you are
log_indiv_quality <- rnorm(n = nfly, mean = 0, sd = .2)

exp(log_avg_larva_time)

exp(log_avg_larva_time + log_to_pupa)


fake_flies_intercept <- tibble(
  fly_id = 1:570,
  fly_code = paste0("fly", fly_id)
) |> 
  rowwise() |> 
  mutate(lh = list(data.frame(stage_id = c(0,1),
                              stage_nm = c("larv", "pupa"))))

true_shape_intercept <- 72

fake_intercept_obs <- fake_flies_intercept |> 
  unnest(cols = c("lh")) |> 
  mutate(avg_duration = log_avg_larva_time + log_to_pupa*stage_id,
         ecart_indiv = log_indiv_quality[fly_id],
         fly_duration = avg_duration + ecart_indiv) |> 
  rowwise() |> 
  mutate(obs_duration = rgamma(1, true_shape_intercept, rate = true_shape_intercept/exp(fly_duration)))

fake_intercept_obs |> 
  ggplot(aes(x = obs_duration, fill = stage_nm)) + geom_histogram()




```

```{r}

fake_intercept_obs |> 
  select(fly_id, stage_nm, obs_duration) |> 
  ungroup() |> 
  pivot_wider(names_from = stage_nm, values_from = obs_duration) |> 
  ggplot(aes(x = larv, y = pupa)) + geom_point()

```

model to recover

```{r}
gamma_form_intercept <- bf(obs_duration ~ stage_nm + (1| fly_code),
                                   family = brmsfamily("Gamma",
                                                       link = "log",
                                                       link_shape = "log")) + 
  lf(shape ~ 1)


get_prior(gamma_form_intercept, data = fake_intercept_obs)


intercept_priors <- c(
  prior(normal(0,1), class = "b"),
  prior(normal(2.5,1), class = "Intercept"),
  prior(exponential(2), class = "sd"),
  prior(normal(4,2), class = "Intercept", dpar = "shape")
)


intercept_stages_gamma <- brm(gamma_form_intercept,
                               data = fake_intercept_obs, 
                               prior = intercept_priors,
                               file = here::here("_posts", "2022-06-06-hybrid-fly-life-history", "intercept_gamma.rds"),
                               cores = 4)


```

## Genetic lines

* avg for one genotype
* average for the second
* diff between their average and the hybrid

H, N, and hybrid

We can generate this with Helmert contrasts

```{r}
three_helmerts_default <- contr.helmert(3)
three_helmerts <- sweep(cbind(1, three_helmerts_default), MARGIN = 2, STATS = 1:3, FUN = `/`)
three_helmerts

solve(three_helmerts)
```


We have effects and we want to make group differences out of them:

```{r}
# units in days
grand_mean <- 7
# n develops 2 days slower
h_n_contr <- 2
# hybrids develop 5 days slower 
hybrid_effect <- 0

geno_names <- c("H", "N", "HN")

rownames(three_helmerts) <- geno_names

avg_effects <- tibble(geno = rep(geno_names, each = 20),
       geno_mean = grand_mean*three_helmerts[geno, 1] + h_n_contr*three_helmerts[geno, 2] + hybrid_effect*three_helmerts[geno, 3])

avg_effects |> 
  ggplot(aes(x = factor(geno, levels = geno_names), y = geno_mean)) + geom_point()

distinct(avg_effects)

```

This also works on the log scale:

```{r}
# units in days
log_grand_mean <- log(7)
# n develops 30% longer
log_h_n_contr <- log(1.3)
# hybrids develop 10% longer than expected 
log_hybrid_effect <- log(1.1)

avg_effects <- tibble(geno = rep(geno_names, each = 20),
       geno_mean = log_grand_mean*three_helmerts[geno, 1] + 
         log_h_n_contr*three_helmerts[geno, 2] +
         log_hybrid_effect*three_helmerts[geno, 3])

avg_effects |> 
  ggplot(aes(x = factor(geno, levels = geno_names), y = exp(geno_mean))) + 
  geom_point()

distinct(avg_effects) |> 
  mutate(exp(geno_mean))
```

### extend this to variation effects


```{r}
# reminder of how the gamma works

rr <- rgamma(3000, shape = exp(4), rate = exp(4)/30)
hist(rr)
mean(rr)
var(rr)
30^2/exp(4)

sd(rr)

30/sqrt(exp(4))


curve(1/sqrt(exp(x)), xlim = c(-3, 6))

```


Use parameterization of gamma used by brms: the parameter $\alpha$ scales the square of the variance to the mean.

```{r}
# start with shape = 1 for simplicity!

log_shape <- 6
library(ggridges)
avg_effects |> 
  mutate(obs_time = rgamma(nrow(avg_effects), shape = exp(log_shape), rate = exp(log_shape)/exp(geno_mean))) |> 
  ggplot(aes(y = geno, x = obs_time)) + 
  stat_density_ridges() + 
  coord_cartesian(xlim = c(2, 10))
  
  
```

## now let genotype control the variance AND the average

```{r}

shape_params <- c(grand_var = log(9), diff_h_n = log(1), hybrid_effect = log(.7))
three_helmerts %*% shape_params
exp(three_helmerts %*% shape_params)

avg_var_effects <- avg_effects |> 
  # add shape AKA concentration parameters
  mutate(geno_shape = shape_params[1]*three_helmerts[geno, 1] + 
           shape_params[2]*three_helmerts[geno, 2] +
           shape_params[3]*three_helmerts[geno, 3])

time_observations_onestage <- avg_var_effects |> 
  mutate(obs_time = rgamma(nrow(avg_var_effects), shape = exp(geno_shape), rate = exp(geno_shape)/exp(geno_mean)))


time_observations_onestage |> 
  ggplot(aes(x = obs_time, y = geno)) + stat_density_ridges()

```




```{r eval=FALSE}


nfly <- 570
# average time you spend as a larva
log_avg_larva_time <-  log(13)
# average time DIFFERECE to the time you spend as a pupa (might be close to 0)
log_to_pupa <- -.2
# your "quality", some idea of how much above and below the population you are
log_indiv_quality <- rnorm(n = nfly, mean = 0, sd = .2)

exp(log_avg_larva_time)

exp(log_avg_larva_time + log_to_pupa)

genotype_vals <- tibble(genotype = rep(c("H", "N", "HN"), each = 100),
                        fly_id = 1:300) |> 
  rowwise() |> 
  mutate(lh = list(data.frame(stage_id = c(0,1),
                              stage_nm = c("larv", "pupa")))) |> 
  unnest(cols ="lh")

true_shape_intercept <- 72

genotype_vals |> 
  rowwise() |> 
  mutate(avg_duration = 
           log_avg_larva_time + 
           log_to_pupa*stage_id +  
           log_indiv_quality[fly_id] + genotype_mean[genotype])


```




