---
title: "FID: When to run away"
description: |
  Modelling a decay to a constant
author:
  - name: Catherine Capkun-Huot and Andrew MacDonald
date: 05-26-2022
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r eval = FALSE}
form_fit4 <- bf(FID    ~ inv_logit(logitM) * 1000 * (1 - inv_logit(logitp)*num_obs/(exp(logd) + num_obs)),
                logitM ~ 1 + risk + scale(capt) + (1 |t| ID),
                logitp ~ 1 + risk + sex + scale(docil) + scale(explo) + (1 |t| ID),
                logd   ~ 1 + risk + (1 |t| ID),
                nl     = TRUE,
                family = Gamma(link = "identity"))
```


* i for observations (obs, exposure)
* j for individuals / groups / units (individual chipmunks)
* k for risk
* 
* k, l, m etc for others

$$
\begin{align}
\text{FID}_{ijk} \sim& \text{gamma}\left(\alpha, \frac{\alpha}{\mu_{ijk}}\right)\\
\mu_{ijk} =& m_{jk} \times 1000 \times \left( 1 - \frac{p_{jk} \times X_i}{d_{jk} + X_i}  \right) \\
\text{logit}(m_{jk}) =& \bar{m} + \beta_k^m + m_{jk}\\
\text{logit}(p_{jk}) =& \bar{p} + \beta_k^p + p_{jk}\\
\text{log}(d_{jk})   =& \bar{d} + \beta_k^d + d_{jk}\\
\begin{bmatrix}
m_{k}\\
p_{k}\\
d_{k}\\
\end{bmatrix} \sim& \text{multivariate normal}\left(\begin{bmatrix}0\\0\\0\end{bmatrix}, \text{diag}(\sigma)\times \textbf{R}\times\text{diag}(\sigma)\right) \\ 
\bar{m}, \beta_k^m \sim& \text{normal}(.5, .5)\\

\bar{p},\beta_k^p \sim& \text{normal}(-1, .2)\\

\bar{d},\beta_k^d \sim& \text{normal}(1.5, .5)\\
\alpha \sim& \text{gamma}(6.25, .25) \\ 
\textbf{R} \sim&\text{LKJ}(2) \\
\sigma \sim& \text{exponential}(2) \\
\beta_k
\end{align}
$$


All models were fit with the following priors: the fixed effects had normal distributions (logitM: mean = 0.5, sd = 0.5; logitp: mean = – 1, sd = 0.2; logd: mean = 1.5, sd = 0.5), the random effects had exponential distributions (lamda = 2), the shape parameter had a gamma distribution (alpha = 6.25, beta = 0.25)[CCH1] , and the covariance matrix had a Lewandowski-Kurowicka-Joe (LKJ) distribution (eta = 2) .





---------------------

$$
\begin{align}
\text{FID}_i \sim& \text{Gamma}\left(\frac{\mu_i}{\sigma^2}, \frac{\mu_i^2}{\sigma^2}\right)\\
\mu_i =& m_i \times 1000 \times \left( 1 - \frac{p_i \times \text{Obs}}{d_i + \text{Obs}}  \right) \\
\text{logit}(m_i) =& \bar{m} + \beta_{m,\text{risk}[i]} + m_{\text{tamia}[i], \text{risk}[i]}\\
\text{logit}(p_i) =& \bar{p} + \beta_{p,\text{risk}[i]} + p_{\text{tamia}[i], \text{risk}[i]}\\
\text{log}(d_i)   =& \bar{d} + \beta_{d,\text{risk}[i]} + d_{\text{tamia}[i], \text{risk}[i]}\\
\begin{bmatrix}
m_{\text{tamia}, \text{risk}[i]}\\
p_{\text{tamia}, \text{risk}[i]}\\
d_{\text{tamia}, \text{risk}[i]}
\end{bmatrix} &\sim \text{MVnormal}\left(\begin{bmatrix}0\\0\\0\end{bmatrix}, \textbf{S}_{\text{risk}[i]}\right)
\end{align}
$$

$j$ indicates the chipmunks present in each treatment, of which there are $J_r$

put real numbers in? why not?

 <!-- \begin{bmatrix}\sigma_{m, \text{risk}[i]}\\\sigma_{p, \text{risk}[i]}\\\sigma_{d, \text{risk}[i]}\end{bmatrix}, R_{\text{risk}[i]}\right -->

```{r}
mygamma <- function(x, a, u){
  (a/u)^a/gamma(a)*x^(a-1)*exp(-a*x/u)
}

curve(mygamma(x, 50, 10), xlim = c(0, 24))

```

